\mythesistitleNOTSYNPOSIS \tnote[fyi]{Synopsis title is ``Generalization of Consecutive Ones Property of Binary Matrices''}

### 2012-01-24 src/020chap2.tex
Did some over all clean up. Added fancy chapter headings. Begun work on opt survey.

\chapter{\CoP\,-- A Survey of Important Results}
\label{ch:copsurvey}

@@

Table~\ref{tab:graphmatrices} summarises the way these graphs are
characterized by their matrices having \COP or \CROP.
Our focus in this chapter (and thesis) is mainly \COP and 
having seen how useful \COP is in identifying or characterizing many
types of graphs\footnote{\COP has several other applications
  \temptext{fill up.}}, we will now see results that study recognition of \COP
in matrices in the following section.

\temptext{ tab~\ref{tab:graphmatrices} -  have an abridged version
  of table 2.1 in dom. see notes for the possibility of a
  ``circle diagram''.} 
\begin{table}[htbp]
  \centering
% \includegraphics[scale=0.1]{../img/circlediagram_domtab2.1__2.png} %
 \caption{\figtabsize Graph matrices \temptext{PLACEHOLDER}}
  \label{tab:graphmatrices}
\end{table}

@@

\tnote[important]{1899 by petrie -- cite
         kendall 69 pacific journal of mathematics 1969 }

\def\tabparboxw{10cm}
\begin{table}[htbp]
  \centering
  {
    \fontencoding{OT1}\fontfamily{cmss}\selectfont%
    \footnotesize%

    \begin{tabular}{lll}
      \hline
      \hline

      1899 & \parbox[t]{\tabparboxw}{ First mention of \COP
        (archaeology)} &                  \\

      \rowcolor[gray]{0.8} %
      1951 & \parbox[t]{\tabparboxw}{ Heuristics for COT} & \cite{wsr51} \\

      1965 & \parbox[t]{\tabparboxw}{ First polynomial time algorithm
        for \COP testing} & \cite{fg65} \\

      \rowcolor[gray]{0.8} %
      1972 & \parbox[t]{\tabparboxw}{ Characterization for \COP -- forbidden
        submatrices} &
      \cite{at72} \\

      1976 & \parbox[t]{\tabparboxw}{ First linear time algorithm for COT --
        \PQtree} &
      \cite{bl76} \\

      \rowcolor[gray]{0.8} %
      1992 & \parbox[t]{\tabparboxw}{ Linear
        time algorithm COT without \PQtree} & \cite{wlh02}%\footnote{First
      % published in \cite{wlh02isaac92}}
      \\

      2001 & \parbox[t]{\tabparboxw}{\PCtree\ -- a simplification of
        \PQtree} & \cite{wlh01,hm03} \\

      \rowcolor[gray]{0.8} %     
      1996 & \parbox[t]{\tabparboxw}{\PQRtree\ -- generalization of
        \PQtree for any binary matrix regardless of its \COP status} &
      \cite{mm96} \\

      1998 & \parbox[t]{\tabparboxw}{ Almost linear time to construct
        \PQRtree} & \cite{mpt98} \\

      \rowcolor[gray]{0.8} %
      2004 & \parbox[t]{\tabparboxw}{ A certifying algorithm for no
        \COP.\\ Generalized \PQtree.} & \cite{mcc04} \\

      2009 & \parbox[t]{\tabparboxw}{ Set theoretic, cardinality based
        characterization of \COP\ -- ICPIA} & \cite{nsnrs09} \\

      \rowcolor[gray]{0.8} %
      2010 & \parbox[t]{\tabparboxw}{ Logspace \COP testing} &
      \cite{kklv10}\\ 
      \hline
      \hline
    \end{tabular}
  }
\caption[\figtabsize A brief history of \COP research]{\figtabsize A
  brief history of \COP research }
\label{tab:cophistory}
\end{table}

@@

\subsection{Tucker's forbidden submatrices for \COP}

% \change{While it is remarkable that the first polynomial time
%   algorithm for \COP testing came in }{The first polynomial time
%   algorithm for \COP testing came in }\cite{fg65} \change{,}{.}
% \change{it is safe to say that the most}{A few years later, a very}
% significant result in understanding \COP came \remove{a few years
%   later} from Tucker which gave a combinatorial (negative)
% characterization of matrices with COP \cite{at72}.

@@

\begin{theorem}
  (\cite[Th.~6]{at72}, \cite[Th.~2.3]{d08phd})\\
  A bipartite graph $G = (V_1, V_2, E)$ is convex bipartite on
  columns\footnote{Abridged to match terminology adopted in this
    document. See previous note.} \iff $V_1$ contains no asteroidal
  triple of $G$.
  \label{th:tuckeratfree}
\end{theorem}

\begin{theorem}
  (\cite[Th.~7]{at72}, \cite[Th.~2.4]{d08phd})\\
  In a bipartite graph $G = (V_1, V_2, E)$ the vertex set $V_1$
  contains no asteroidal triple \iff $G$ contains none of the graphs
  $G_{I_k}$, $G_{II_k}$, $G_{III_k}$ (with $k \ge 1$), $G_{IV}$,
  $G_{V}$ as shown in Figure~\ref{fig:forbiddensubgraphs} as subgraphs.
  \label{th:tuckerforbidden}
\end{theorem}

@@

\begin{theorem}
  (\cite[Th.~9]{at72}, \cite[Th.~2.5]{d08phd})\\
  A matrix $M$ has \COP \iff it contains none of the matrices 
$M_{I_k}$, $M_{II_k}$, $M_{III_k}$ (with $k \ge 1$), $M_{IV}$,
  $M_{V}$ as shown in Figure~\ref{fig:forbiddensubmatrices} as submatrices.
  \label{th:tuckercop}
\end{theorem}

\begin{figure}[htbp]
  \centering
%  \includegraphics[scale=0.35]{../img/tuckersforbiddensubmatrices.png}  
  \setlength\dashlinedash{1pt} %
  \setlength\dashlinegap{0pt}  %  Make non zero if cline can be done
                               %   with dashed lines.
  \singlespacing
  \begin{tabular}[h]{ll}
%   \small 
   \begin{tabular}[t]{|lll ll l|l} %                   MATRIX 1
      \multicolumn{7}{l}{\normalsize \hspace{-0.5in} $M_{I_k}, k \ge 1$}              \\   
      \multicolumn{6}{c}{\matrixbracetop{3.25}{$k+2$}}   \\
      \cline{1-6} % \firsthline %      Begin matrix  %%%%%%
      \un & \un & 0   & \multicolumn{2}{c}{\dots}  & 0    &
        \hspace{-0.1in}\multirow{5}{*}{\matrixbraceright{2.5}{$k+2$}}   \\    
      0   & \un & \un & 0 & ...& 0   \\
      \multicolumn{6}{|c|}{\dots}                        \\
      0   & \multicolumn{2}{c}{\dots}  & 0   & \un & \un \\
      \cline{1-6} % \hdashline %%%%%%%%%%%%%%%%%%%%%%%%%%%%
      \un & 0   & \multicolumn{2}{c}{\dots}  & 0   & \un \\
      \cline{1-6} % \lasthline %      End matrix  %%%%%%%%%
    \end{tabular}
    &%
%    \small
    \begin{tabular}[t]{|lll ll ll|l} %                   MATRIX 2
      \multicolumn{8}{l}{\normalsize \hspace{-0.5in}$M_{II_k}, k \ge 1$}                   \\
      \multicolumn{7}{c}{\matrixbracetop{3.8}{$k+3$}}            \\ 
      \cline{1-7}%  \firsthline %      Begin matrix  %%%%%%%%%%%%
      \un & \un & 0   & \multicolumn{3}{c:}{\dots}       & 0    &
        \hspace{-0.1in}\multirow{6}{*}{\matrixbraceright{3}{$k+3$}}         \\
      0   & \un & \un & 0 & \multicolumn{2}{c:}{\dots}   & 0   \\ 
      \multicolumn{6}{|c:}{\dots}                        &     \\
      0   & \multicolumn{2}{c}{\dots}  & 0   & \un              &
        \multicolumn{1}{l:}{\un} & 0                           \\
      \cline{1-6}    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      0   & \un & \multicolumn{4}{c}{\dots}              & \un \\
      \un & \multicolumn{3}{c}{\dots}        & \un & 0   & \un \\
      \cline{1-7}% \lasthline %      End matrix  %%%%%%%%%%%%%%%%
    \end{tabular}
    \\\\%
%    \small
    \begin{tabular}[t]{|lll ll l l|l} %                  MATRIX 3 
      \multicolumn{8}{l}{\normalsize \hspace{-0.5in}$M_{III_k}, k \ge 1$}                  \\
      \multicolumn{7}{c}{\matrixbracetop{3.8}{$k+3$}}            \\
      \cline{1-7}%      \firsthline %      Begin matrix %%%%%%%%%
      \un & \un & 0   & \multicolumn{3}{c:}{\dots}       & 0    &
        \hspace{-0.1in}\multirow{5}{*}{\matrixbraceright{2.5}{$k+2$}}         \\
      0   & \un & \un & 0 & \multicolumn{2}{c:}{\dots}   & 0   \\ 
      \multicolumn{6}{|c:}{\dots}                        &     \\
      0   & \multicolumn{2}{c}{\dots}  & 0   & \un              &
        \multicolumn{1}{l:}{\un}                         & 0   \\
      \cline{1-6}    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
      0   & \un & \multicolumn{2}{c}{\dots}  & \un & 0   & \un \\
      \cline{1-7}%      \lasthline %      End matrix %%%%%%%%%%%%
    \end{tabular}
    &%
%    \small
    \begin{tabular}[t]{|llllll|}  %                     MATRIX 4
      \multicolumn{6}{l}{\normalsize \hspace{-0.5in}$M_{IV}$} \\
      \multicolumn{6}{l}{} \\
      \firsthline
      \un & \un & 0 & 0 & 0 & 0   \\
      0 & 0 & \un & \un & 0 & 0   \\
      0 & 0 & 0 & 0 & \un & \un   \\
      \un & 0 & \un & 0 & \un & 0 \\
      \lasthline
    \end{tabular}
    \\\\%
%    \small
    \begin{tabular}[t]{|lllll|}  %                     MATRIX 5
      \multicolumn{5}{l}{\normalsize \hspace{-0.5in}$M_{V}$} \\
      \multicolumn{5}{l}{} \\
      \firsthline
      \un & \un & 0   & 0   & 0   \\
      0   & 0   & \un & \un & 0   \\
      \un & \un & \un & \un & 0   \\
      \un & 0   & \un & 0   & \un \\
      \lasthline
    \end{tabular}
  \end{tabular}

  \caption[\figtabsize Tucker's forbidden submatrices]{\figtabsize Tucker's
    forbidden submatrices for convex bipartite graphs. \cite{at72}} 
  \label{fig:forbiddensubmatrices}
\end{figure}

@@

\begin{definition}[{\em \Cop for set systems}]%
  \label{def:matrixsetsystem}%
  % [\emph{Set system of a binary matrix}.]
  Let $M$ be a binary matrix of order $n \times m$ and $\{c_i \mid i
  \in [m]\}$ be the columns in $M$.  A set system $\cF_{M} = \{S_i
  \mid S_i \subseteq [n], i \in [m]\}$ is defined such that for every
  column $c_i$ of $M$, set $S_i = \{j \mid m_{ji} = \un \}$. The
  collection $\cF_M$ is the {\em set system of binary matrix} $M$. The
  {\em binary matrix for set system} $\cF$ is conversely constructed
  and denoted by $M^\cF$. Thus, $M^{\cF_M} = M$.

  A set system $\cF$ from universe $U$, $|U| = n$ has the
  {\em \cop} if there exists a linear order or 
  permutation $\sigma = w_1w_2\ldots w_n$ that can be applied to $U$ \stt each set $S \in
  \cF$ becomes a consecutive subsequence\footnote{Also termed an {\em interval}} $w_{i}w_{i+1}\ldots
  w_{i+k-1}$ on $\sigma$ for some positive integer $i \le n+1-k$ where  $k = |S|$.

  The set $\valid{U,\cF}$ respresents all \COP orders of $\cF$ in $U$.
  \defnstop
\end{definition}

@@

  \item $\overline{\cF}$ represents the smallest super set system of
    $\cF$ that is complete\footnote{\cite[Def.~3.2]{mcc04} calls
      this the {\em weak closure} of $\cF$ denoted by $\cW(\cF)$. \label{mcc3}}

  \end{enumerate}
  \defnstop
\end{definition}

@@

\begin{theorem}[{\cite[Th.~3,6]{mm96}}]
  \label{th:validcop}
  For any set system $\cF$ we have the following. \par
  \centering
    $\valid{\cF} = \valid{\overline{\cF}}$ \\
    $\cF^{\bot} = \overline{\cF^{\bot}} = (\overline{\cF})^{\bot}$
\end{theorem}

@@

\begin{theorem}[{\cite[Th.~7]{mm96}}]
  \label{th:mmdecomp}
  For any set system $\cF$, and $\emptyset \ne H \in \overline{\cF}
  \cap \cF^{\bot}$ we have the following.\par
  \centering
  $\valid{U,\cF} = \valid{U/H,\cF/H} * \valid{H,\cF \cap 2^H}$
\end{theorem}

@@

We observe that this idea of decomposing the \COP problem into two
subproblems in \cite{mm96} is very similar to the substitution
decomposition of a set system given in \cite{mcc04}.\footnote{Using
  the theory cited in footnotes~\ref{mcc1},\ref{mcc2},\ref{mcc3}.}

@@

\begin{theorem}[{\cite[Th.~9]{mm96},\cite[Th.~2.1]{mcc04}}]
\label{th:pqrfinalmm}
If $\cF$ is a set system with universe $U$ and $|U| \ge 3$ then one of
the following statements hold.
\begin{enumerate}
\item \label{th::it1} $\overline{\cF} = \cT(U)$ 
\item \label{th::it2}$\overline{\cF} = consec(\alpha)$\tnote{what is consec?} for
  some permutation $\alpha$ on $U$ 
\item \label{th::it3}$\overline{\cF} = \power{U}$ 
\end{enumerate}
\end{theorem}

@@

\begin{theorem}[{\cite{mm96}, \cite[Th.~3.6]{mcc04}}]
The set system $\cF$ has \COP \iff its \PQRtree has no $R$ nodes.  
\end{theorem}

% The generalized \PQtree of $\cF$ is defined as the {\em decomposition
%   tree} of its weak closure $\cW(\cF)$. Decomposition tree is the
% inclusion tree (inclusion p.o. Hasse diagram) of the strong elements
% of $\cW(\cF)$ with labels {\em prime, linear} and {\em degenerate}.\cite{mcc04}

% Overlap components are computable in linear time \cite{mm95,wlh92}.

----- Cert algo

------ ICPIA

\temptext{ Certifying algorithm (Generalized PQ trees) [McC04] --

Set theoretic characterizations [Hsu02, NS09]\\
    \cite{wlh02} describes the simpler algorithm for
    COT.\tnote[pressing]{in terms of what?}\\
    \cite{nsnrs09} describes a characterization of consecutive-ones
    property solely based on the cardinality properties of the set
    representations of the columns (rows); every column (row) is
    equivalent to a set that has the row (column) indices of the rows
    (columns) that have one entries in this column (row). This is
    interesting and relevant, especially to this thesis because it
    simplifies COT to a great degree by reducing the solution search
    space owing to the a simple set theoretic characterization.\\

\cite{mcc04} describes a different approach to COT. While all previous
COT algorithms gave the COP order if the matrix has the property but
exited stating negative if otherwise, this algorithm gives an evidence
by way of a certificate of matrix even when it has no COP. This
enables a user to verify the algorithm's result even when the answer
is negative. This is significant from an implementation perspective
because automated program verification is hard and manual verification
is more viable. Hence having a certificate reinforces an
implementation's credibility. Note that when the matrix {\em has} COP,
the COP order is the certificate.  The internal machinery of this
algorithm is related to the weighted betweenness problem
addressed\tnote{in what way??} in \cite{co98}.  \tnote{expand
  on the COP order graph creation and it having to be bipartite for M
  to have COP. and thus an odd cycle being an evidence of no COP.}

@@

\PCtree is a generalization of \PQtree. It is a data structure that is
analogous to \PQtree but for matrices with \crop. \PCtree was
introduced by \cite{sh99} for the purpose of planarity testing where
this data structure represents partial embeddings of planar
graphs. % noted from wlh01 abstract.
In \cite{wlh01}, Hsu reintroduces \PCtree as a generalization of
\PQtree and shows how it simplifies \cite{bl76}'s planarity test by
making the \PQtree construction much less complicated. Later
\cite{hm03}, discovers that \PCtree is a representation of all \crop
orders of a matrix when it is unrooted. \PCtree presented in
\cite{wlh01} is rooted; however the construction of \PCtree is the
same in both results. The property of being unrooted is necessary in
order to use \PCtree as a data structure for encoding circular
ordering. Definition~\ref{def:pctree} defines \PCtree.

@@

\section{Optimization problems in COP}
\label{sec:surveycopopt}
\tnote{Expand on ref:sec:optcop}

>>>>

%\chapter{\CoP\,-- A Survey of Important Results}
\chapter{Consecutive-ones Property -- A Survey of Important Results}
\label{ch:copsurvey}

@@

Table~\ref{tab:graphmatrices} summarises the way these graphs are
characterized by their matrices having \COP or \CROP.
Our focus in this chapter (and thesis) is mainly \COP and 
having seen how useful \COP is in identifying or characterizing many
types of graphs, we will now see results that study recognition of \COP
in matrices in the following section.

\tabgraphmatrices

@@

\tabcophistory

@@

\subsection{Tucker's forbidden submatrices for \COP}
\label{sec:tucker}
% \change{While it is remarkable that the first polynomial time
%   algorithm for \COP testing came in }{The first polynomial time
%   algorithm for \COP testing came in }\cite{fg65} \change{,}{.}
% \change{it is safe to say that the most}{A few years later, a very}
% significant result in understanding \COP came \remove{a few years
%   later} from Tucker which gave a combinatorial (negative)
% characterization of matrices with COP \cite{at72}.

@@

\begin{theoremsansproof}
  [{\cite[Th.~6]{at72}, \cite[Th.~2.3]{d08phd}}]%\\
  A bipartite graph $G = (V_1, V_2, E)$ is convex bipartite on
  columns\footnote{Abridged to match terminology adopted in this
    document. See previous note.} \iff $V_1$ contains no asteroidal
  triple of $G$.
  \label{th:tuckeratfree}
\end{theoremsansproof}

\begin{theoremsansproof}
  [{\cite[Th.~7]{at72}, \cite[Th.~2.4]{d08phd}}]%\\
  In a bipartite graph $G = (V_1, V_2, E)$ the vertex set $V_1$
  contains no asteroidal triple \iff $G$ contains none of the graphs
  $G_{I_k}$, $G_{II_k}$, $G_{III_k}$ (with $k \ge 1$), $G_{IV}$,
  $G_{V}$ as shown in Figure~\ref{fig:forbiddensubgraphs} as subgraphs.
  \label{th:tuckerforbidden}
\end{theoremsansproof}

@@

\begin{theoremsansproof}
  [{\cite[Th.~9]{at72}, \cite[Th.~2.5]{d08phd}}]%\\
  A matrix $M$ has \COP \iff it contains none of the matrices 
$M_{I_k}$, $M_{II_k}$, $M_{III_k}$ (with $k \ge 1$), $M_{IV}$,
  $M_{V}$ as shown in Figure~\ref{fig:forbiddensubmatrices} as submatrices.
  \label{th:tuckercop}
\end{theoremsansproof}

\figforbiddensubmatrices

@@

\begin{definition}[{\em \Cop for set systems}]%
  \label{def:matrixsetsystem}%
  % [\emph{Set system of a binary matrix}.]
  Let $M$ be a binary matrix of order $n \times m$ and $\{c_i \mid i
  \in [m]\}$ be the columns in $M$.  A set system $\cF_{M} = \{S_i
  \mid S_i \subseteq [n], i \in [m]\}$ is defined such that for every
  column $c_i$ of $M$, set $S_i = \{j \mid m_{ji} = \un \}$. The
  collection $\cF_M$ is the {\em set system of binary matrix} $M$. The
  {\em binary matrix for set system} $\cF$ is conversely constructed
  and denoted by $M^\cF$. Thus, $M^{\cF_M} = M$.%
  \par\noindent%
  A set system $\cF$ from universe $U$, $|U| = n$ has the
  {\em \cop} if there exists a linear order or 
  permutation $\sigma = w_1w_2\ldots w_n$ that can be applied to $U$ \stt each set $S \in
  \cF$ becomes a consecutive subsequence\footnote{Also termed an {\em interval}} $w_{i}w_{i+1}\ldots
  w_{i+k-1}$ on $\sigma$ for some positive integer $i \le n+1-k$ where  $k = |S|$.
  \par\noindent%
  The set $\valid{U,\cF}$ respresents all \COP orders of $\cF$ in $U$.
  \dstop
\end{definition}

@@

  \item $\overline{\cF}$ represents the smallest super set system of
    $\cF$ that is complete\footnote{\cite[Def.~3.2]{mcc04} calls
      this the {\em weak closure} of $\cF$ denoted by $\cW(\cF)$. \label{mcc3}}
  \end{enumerate}
  \dstop
\end{definition}

@@

\begin{theoremsansproof}[{\cite[Th.~3,6]{mm96}}]
  \label{th:validcop}
  For any set system $\cF$ we have the following. \par
  \centering
    $\valid{\cF} = \valid{\overline{\cF}}$ \\
    $\cF^{\bot} = \overline{\cF^{\bot}} = (\overline{\cF})^{\bot}$
\end{theoremsansproof}

@@

\begin{theoremsansproof}[{\cite[Th.~7]{mm96}}]
  \label{th:mmdecomp}
  For any set system $\cF$, and $\emptyset \ne H \in \overline{\cF}
  \cap \cF^{\bot}$ we have the following.\par
  \centering
  $\valid{U,\cF} = \valid{U/H,\cF/H} * \valid{H,\cF \cap 2^H}$
\end{theoremsansproof}

@@

We observe that this idea of decomposing the \COP problem into two
subproblems in \cite{mm96} is very similar to the substitution
decomposition of a set system given in \cite[Sec.~4]{mcc04}.\footnote{Using
  the theory cited in footnotes~\ref{mcc1},\ref{mcc2},\ref{mcc3}.}

@@

\begin{theoremsansproof}[{\cite[Th.~9]{mm96},\cite[Th.~2.1]{mcc04}}]
\label{th:pqrfinalmm}
If $\cF$ is a set system with universe $U$ and $|U| \ge 3$ then one of
the following statements hold.
\begin{enumerate}
\item \label{th::it1} $\overline{\cF} = \cT(U)$ 
\item \label{th::it2}$\overline{\cF} = consec(\alpha)$\tnote{what is consec?} for
  some permutation $\alpha$ on $U$ 
\item \label{th::it3}$\overline{\cF} = \power{U}$ 
\end{enumerate}
\end{theoremsansproof}

@@

\begin{theoremsansproof}[{\cite{mm96}, \cite[Th.~3.6]{mcc04}}]
The set system $\cF$ has \COP \iff its \PQRtree has no $R$ nodes.  
\end{theoremsansproof}

Thus \PQRtree gives a data structure that encodes possible linear
orderings of a set that demonstrates the \COP property in it or
narrows it down to parts of the universe in $R$ nodes that prevent the
set system from having \COP.

As mentioned before, we observe the three data strutures of \PQRtree,
\gPQtree and generalized \PQtree to be equivalentand the theory of all
these data structures is summarised in Table~\ref{tab:pqrcomparison}.

We will now see one more generalization of \PQtree before seeing other
approaches to solving \COP testing including ICPIA which is a set
theoretic characterization of \COP.

\begin{table}[t]
  \def\colwidth{4cm}
  \centering
  \fontencoding{OT1}\fontfamily{cmss}\selectfont%
  \footnotesize%
  \setlength\extrarowheight{0.1in}

  \begin{tabular}{l >{\columncolor{\tblhcolor}}l l}

    \rowcolor[gray]{0.8} %                 Header row colored
    \normalsize \gPQtree \cite{n89} & 
    \normalsize \PQRtree \cite{mm96} & 
    \normalsize generalized \PQtree \cite{mcc04}
    \\

    Trivial intersections &
    Trivial sets $\cT(U)$ & 
    \\\hline

    Trivially intersecting sets &
    Mutually orthogonal sets, $A \bot B$ &
    Non-overlapping sets
    \\\hline

      &
    Complete collection &
    Weakly partative family
    \\\hline

    &
    $\overline{\cF}$&
    Weak closure, $\cW(\cF)$
    \\\hline

    &
    $\cF^{\bot}$&
    Non-overlapping family, $\cN(\cF)$
    \\\hline

    &
    \parbox[t]{\colwidth}
    { 
      Sets in $\cF$ orthogonal to all other sets in $\cF$ 
    }&
    Strong elements
    \\\hline

    &
%     \parbox[t]{\colwidth}
%       {}
    \temptext{PQR tree def. }
    &
    \parbox[t]{\colwidth}
    {
      Decomposition tree of $\cW(\cF)$, $T(\cW(\cF))$
      \temptext{PQR tree def. }\tnote[pressing]{in terms of theory}
    }
    \\\hline

    &
    Node sets, $(\cF \cap \cF^\bot) \setminus \{\emptyset\}$
    \\\hline

    &
    \parbox[t]{\colwidth}
    {
      $P$-node $\Leftrightarrow \cF = \cT(\cF)$\\
      $Q$-node $\Leftrightarrow \cF = consec(\alpha)$\\
      $R$-node $\Leftrightarrow \cF = \power{U}$\\
    }&
    \temptext{give def of P Q R nodes}
    \\\lasthline

  \end{tabular}
  \caption[\figtabsize Comparison of theory of \PQRtree, \gPQtree, generalized
  \PQtree]{\figtabsize Comparison of theory of \cite{mm96} \PQRtree, \cite{n89} \gPQtree,
    \cite{mcc04} generalized \PQtree}
  \label{tab:pqrcomparison}
\end{table}

@@

\PCtree is another generalization of \PQtree. It is a data structure that is
analogous to \PQtree but for matrices with \crop. \PCtree was
introduced by \cite{sh99} for the purpose of planarity testing where
this data structure represents partial embeddings of planar
graphs. % noted from wlh01 abstract.
In \cite{wlh01}, Hsu reintroduces \PCtree as a generalization of
\PQtree and shows how it simplifies \cite{bl76}'s planarity test by
making the \PQtree construction much less complicated. Later
\cite{hm03}, discovers that \PCtree is a representation of all \crop
orders of a matrix when it is unrooted. \PCtree presented in
\cite{wlh01} is rooted; however the construction of \PCtree is the
same in both results. The property of being unrooted is necessary in
order to use \PCtree as a data structure for encoding circular
ordering. Definition~\ref{def:pctree} defines \PCtree.

@@

\subsection{ICPIA - a set cardinality based \COP test}

\cite{nsnrs09} describes a characterization of consecutive-ones
property solely based on the cardinality properties of the sets in the
set system and does not use any variants of \PQtrees.

\begin{definition}[Intersection Cardinality Preserving Interval
  Assignment (ICPIA)]
  \label{def:icpia}
  Let $\cF = \{A_i \mid A_i \subseteq U, i \in [m]\}$ be a set system
  from universe $U$ and the set of ordered pairs $\Pi = \{(A_i, B_i)
  \mid B_i  \text{ is an interval from } [n], i \in [m]\}$ be an {\em interval
    assigment} of $\cF$, then it is called an {\em ICPIA} if it has
  the following properties.
  \begin{enumerate}[i.]
  \item $|A_i| = |B_i|$ for all $i \in [m]$
  \item $|A_i \cap A_j| = |B_i \cap B_j|$ for all $i, j \in [m]$
  \end{enumerate}
\end{definition}

\begin{theoremsansproof}[{\cite[Th.~1]{nsnrs09}}]
  \label{th:icpia1}
  If an interval assignment $\Pi$ is feasible, then it is an ICPIA.
\end{theoremsansproof}

The necessity of ICPIA for a set system to have \COP given in
Theorem~\ref{th:inpia1} is fairly obvious. It turns out that it is
sufficient too.  This is demostrated by two algorithms which work as
follows. The first one iterates over the assignment
$\Pi$ %set system $\cF$
and changes it in each iteration till the altered
assignment %set system
has no more overlapping intervals\footnote{Intervals are the second
  element in each ordered pair of $\Pi$. Def.~\ref{def:icpia}} %sets
\cite[Alg.~1]{nsnrs09}. This is acheived by replacing two overlapping
assignment pairs $(A_i, B_i)$ and $(A_j, B_j)$ with the partitions of
$(A_i \cup A_j,B_i \cup B_j)$ induced by their overlaps -- $(A_i
\setminus A_j, B_i \setminus B_j)$, $(A_i \cap A_j, B_i \cap B_j)$ and
$(A_j \setminus A_i, B_j \setminus B_i)$.

To see it in terms of the theory explained in
Section~\ref{sec:surveycertalgo}, this results in the weak closure of
$\cF$. Let $\cF$, $\cI$ be the set system and interval
system\footnote{A set system where all the sets are intervals.},
respectively, involved in the assignment $\Pi$ and $\cF'$, $\cI'$ be
the same for the output of the algorithm $\Pi'$.  It can be observed
that $\cF' \cup \cF$ is $\overline{\cF}$ or the weak closure of
$\cF$. The analog is true for $\cI \cup \cI'$.  Moreover, this output
$\cF'$, $\cI'$ contains the strong elements of $\overline{\cF}$,
$\overline{\cI}$ respectively. It turns out that $\cI'$ is also an
interval system and that $\Pi'$ is an ICPIA \cite[Lem.~2]{nsnrs09}.

The next algorithm \cite[Alg.~2]{nsnrs09} further refines $\Pi'$ to
$\Pi''$ which represents the family of permutations that yield the
\COP orders represented by $\Pi$ of $\cF$. The subset relation Hasse
diagram of $I'$ is created with each node representing the
corresponding assignment pair of the interval from $\Pi'$. Since all intervals in
$\Pi'$ are non-overlapping, this is a tree. Notice that this means any
two intervals are now either disjoint or one is contained in the
other. Hence this tree is called a {\em containment tree}.  The
algorithm traverses this tree in post order fashion where the post
order function is to replace an assignment pair $(X,Y)$  in $\Pi'$ by their child
assignment pairs in the containment tree, say $(X_1, Y_1),(X_2, Y_2)
\ldots (X_k,Y_k)$. The result $\Pi''$ gives the
\COP order for $\Pi$. These two algorithms prove that ICPIA is a
sufficient condition for the feasibility of the interval assignment.

\begin{theoremsansproof}[{\cite[Th.~2]{nsnrs09}}]
  \label{th:icpia2}
  Let  $\cF = \{A_i \mid A_i \subseteq U, i \in [m]\}$ be a set system
  from universe $U$, $|U| = n$
  and $\Pi = \{(A_i,B_i) \mid B_i \text{ is an interval from } [n], i
  \in [m]\}$ be an ICPIA. Then there exists a permutation $\sigma: [n]
  \rightarrow [n]$ \stt $\sigma(A_i) = B_i$.
\end{theoremsansproof}

What we saw so far in this section is checking for the feasibility of
a given interval assignment. ICPIA can also be used to find an
interval assignment when only the set system $\cF$ is given. Part of
the approach is similar to the overlap component idea in \cite{fg65,
  wlh02}. Once the overlap components are factored out, they can be
independently assigned a subinterval from to which the sets in the
component need to be mapped. This is simple because the overlap
components are disjoint from each other. Now the subproblem is finding
\COP order for each overlap component. In each component, the set
whose intersections with the rest of the sets in that component forms
a single inclusion chain is chosen first and assigned the
leftmost\footnote{By convention. Rightmost can also be chosen
  analogously to get a different \COP order.} interval. The next
candidate set is chosen from this set's overlapping sets. The interval
assigned is calculated by simple intersection cardinality
means\footnote{Assume that the next set overlaps on one of the
  ``ends'' of the current set and calculate the interval by shifting
  left or right of the current interval and adjusting its size to
  match the next set's cardinality.}. Once the next set is assigned an
interval, the current assigned sets are all checked for ICPIA. This is
a backtracking algorithm -- if at any point ICPIA fails, another
overlapping set is chosen. If all overlapping set attempts fail ICPIA,
a different ``first set'' is chosen. This is a remarkably simple
algorithm due to its obvious simplicity in implementation.

\subsection{Other \COP testing algorithms}
There are more \COP test algorithms in the literature. We make a
mention of one of them here.  A simple linear time algorithm was
presented by \cite{hmpv00}. They use something called {\em Lex-BFS
  ordering} of the vertices of a graph to decide in linear time
whether the graph is an interval graph. They also show how any
interval graph recognition algorithm can be used to recognize matrices
with \COP. This is shown in Theorem~\ref{th:intgraphcop}.

\begin{theoremsansproof}[{\cite[Th.~2]{hmpv00}}]
  \label{th:intgraphcop}
  For a binary matrix $M$ the following statements are equivalent.
  \begin{enumerate}
  \item The row adjacency graph $G_r(M)$ is an interval graph and $M$
    is its maximal clique matrix.
  \item The columns of $M$ are maximal and $M$ has the \COP for rows.
  \end{enumerate}
\end{theoremsansproof}

\section{Optimization problems in COP}
\label{sec:surveycopopt}

\COP is a very beneficial property since it simplifies the structure
of the input\footnote{Binary matrix or set system.} leading to
efficient algorithms to otherwise hard problems.  While
Section~\ref{sec:surveycoptest} discusses matrices with \COP and how
to compute \COP orders, this section describes the problems that are
of interest when a matrix does not have this property.  The central
questions in this area are (i) how close is the matrix to having \COP,
mainly in terms of Tucker's forbidden submatrices, and (ii) how
optimally can one alter the matrix to attain \COP.  With regard to the
latter question, recent literature indicate that there has been a lot
of interest in matrix modification problems to make a matrix have \COP
\cite{hg02, tz07}. However, to tackle this problem it is more than
insightful to solve question (i).

To study the smallest hamming distance \temptext{(?)} between the
input matrix and one with \COP, identifying the Tucker's forbidden
substructures\footnote{See Section~\ref{sec:tucker} for detailed
  discussion on Tucker's forbidden substructures for matrices with
  \COP.} is of prime importance.

\subsection{Finding forbidden submatrices}
It is known from Theorem~\ref{th:tuckercop}
that when a matrix does not have \COP, it must have one of Tucker's
forbidden submatrices. However, none of the \COP test algorithms discussed in
Section~\ref{sec:surveycoptest} outputs the forbidden submatrix when
the test is negative. This certainly gives a hint to how hard it is to find
these substructures in the input. \temptext{Fill neg results on
  finding forbidden submartrices -- boo75, some highlighted stuff from v85}

\temptext{TEXT INTRODUCING SECTION}

\subsection{Incompatibility graph - a certificate for no \COP}

When a \COP test algorithm reports negative, it is hard to verify
this. While \cite{at72} gives forbidden substructures for \COP, there
are no efficient algorithms to find these strutures in a matrix. In
fact, \cite{v82} shows that even finding a forbidden submatrix with at
least $k$ rows in an input matrix is NP-complete. This was proven by
polynomially reducing the longest path problem for graphs to this
problem.

\cite{mcc04} describes another characterization for matrices without
\COP which is efficiently computable. This is done by way of an {\em
  incompatibility graph} of a set system\footnote{Since \COP in
  matrices and set systems are equivalent problems as seen in
  Section~\ref{sec:surveycertalgo} we will use the term set system
  here.}. The construction of this graph is based on the following
observation.

\begin{observation} 
  Consider elements $a, b, c \in U$ and a set system $\cF$ with
  universe $U$. If there is at least one set $S \in \cF$ such that $a,
  c \in \cF$ but $b \notin S$, then it is impossible to have a \COP
  order that will place $b$ in between $a$ and $c$. \tnote{also cite
    the inbetweenness problem}
\end{observation}
